Typed Lua: An Optional Type System for Lua

### 2014-03-14 doc/paper/typedlua.tex
started rewrite of related work section

\begin{verbatim}
    local Shape = { x = 0, y = 0 }
    const function Shape:new(x: number, y: number)
      local s = setmetatable({},
                             { __index = self })
      s.x = x
      s.y = y
      return shape
    end
    const function Shape:move(dx: number,
                              dy: number)
      self.x = self.x + dx
      self.y = self.y + dy
    end
    local shape1 = Shape:new(0, 5)
    local shape2 = Shape:new(10, 10)
\end{verbatim}

@@

Notice how we can call {\tt Shape}'s constructor inside
the overriden constructor. A limitation of this class
system is that the overriden constructors must be a
subtype of the original constructor.

@@

We begin this section presenting a little bit of the history
about combining static and dynamic typing in the same language.
After that, we introduce optional type systems and gradual
typing. We end this section discussing why optional type
systems and two other approaches are often called gradual typing.

\subsection{A little bit of history}

Common LISP \citep{steele1982ocl} introduced optional type annotations
in the early eighties, but not for static type checking.
Instead, programmers may choose to declare types of variables as
optimization hints to the compiler, that is, type declarations are
just one way to help the compiler to optimize code.
These annotations are unsafe because they can crash the program
when they are wrong.

\citet{abadi1989dts} extended the simply typed lambda calculus with the
\texttt{Dynamic} type and the \texttt{dynamic} and \texttt{typecase}
constructs, with the aim to safely integrate dynamic code in
statically typed languages.
The \texttt{Dynamic} type is a pair \texttt{(v,T)} where \texttt{v} is a
value and \texttt{T} is the tag that represents the type of \texttt{v}.
The constructs \texttt{dynamic} and \texttt{typecase} are explicit
injection and projection operations, respectively.
That is, \texttt{dynamic} builds values of type \texttt{Dynamic} and
\texttt{typecase} safely inspects the type of a \texttt{Dynamic} value.
Thus, migrating code between dynamic and static type checking requires
changing type annotations and adding or removing \texttt{dynamic} and
\texttt{typecase} constructs throughout the code.

The \textit{quasi-static} type system proposed by \citet{thatte1990qst}
performs implicit coercions and run-time checks to replace the
\texttt{dynamic} and \texttt{typecase} constructs that were proposed by
\citet{abadi1989dts}.
To do that, quasi-static typing relies on subtyping with a top type
$\Omega$ that represents the dynamic type, and splits type checking
into two phases.
The first phase inserts implicit coercions from the dynamic type to
the expected type, while the second phase performs what Thatte calls
\textit{plausibility checking}, that is, it rewrites the program to
guarantee that sequences of upcasts and downcasts always have a
common subtype.

\textit{Soft typing} \citep{cartwright1991soft} is another approach
to combine static and dynamic typing in the same language.
The main goal of soft typing is to add static type checking to
dynamically typed languages without compromising their flexibility.
To do that, soft typing relies on type inference for
translating dynamically typed code to statically typed code.
The type checker inserts run-time checks around inconsistent code and
warns the programmer about the insertion of these run-time checks,
as they mean the existence of potential type errors.
However, the programmer is free to choose between inspecting the
run-time checks and simply running the code.
Soft typing gives this choice to the programmer because it is one of
its key ideas, that is, type inference and static type checking should
not prevent the programmer from running inconsistent code.
One advantage of soft typing is the fact that the compiler for
softly typed languages can use the translated code to generate
more efficient code, as the translated code statically type checks.
One disadvantage of soft typing is that it can be cumbersome when
the inferred types are meaningless large types that just confuse the
programmer.

\textit{Dynamic typing} \citep{henglein1994dts} is an approach
that optimizes code from dynamically typed languages by eliminating
unnecessary checks of tags.
The paper describes how to translate dynamically typed code to
statically typed code that uses a \texttt{Dynamic} type.
The translation is done through a coercion calculus that uses type
inference to insert the operations that are necessary to type check
the \texttt{Dynamic} type during run-time.
Although soft typing and dynamic typing may seem similar, they are not.
They are not similar because soft typing targets statically type
checking of dynamically typed languages for detecting programming
errors while dynamic typing targets the optimization of dynamically
typed code through the elimination of unnecessary run-time checks.
In other words, soft typing sees code optimization as a side effect
that comes with static type checking.

\citet{findler2002chf} proposed contracts for higher-order functions
and blame annotations for run-time checks.
Contracts perform dynamic checking instead of static checking,
but deferring all verifications to run-time can lead to defects
that are difficult to fix, because run-time errors can show a
stack trace where it is not clear to programmers if the cause of a
certain run-time error is in application code or library code.
Even if programmers identify that the source of a certain run-time
error is in library code, they still may have problems to identify
if this run-time error is due to a violation of library's contract,
or due to a bug, when the library is poorly documented.
In this approach, programmers can insert assertions in the form of
contracts that check the input and output of higher-order functions.
Thus, \citet{findler2002chf} show how programmers can insert
assertions in the form of contracts that check the input and output
of higher-order functions; and they also show how to add blame
annotations in the generated code as a way to track assertion
failures back to the source of the error.

BabyJ \citep{anderson2002babyj} is an object-oriented language
without inheritance that allows programmers to incrementally annotate
the code with more specific types.
Programmers can choose between using the dynamically typed version
of BabyJ, when they do not need types at all, and the statically
typed version of BabyJ, when they need to annotate the code.
In the statically typed BabyJ, programmers can use the
\textit{permissive type} $*$ to annotate the parts of the code that
still do not have a specific type or the parts of the code that should
have dynamic behavior.
The type system of BabyJ is nominal, so types are either class names
or the permissive type $*$.
However, the type system does not use type equality or subtyping,
but the relation $\approx$ between two types.
The relation $\approx$ holds when both types have the same name or
any of them is the permissive type $*$.
Even though the permissive type $*$ is similar to the dynamic type
from previous approaches, BabyJ does not provide any way to add
implicit or explicit run-time checks.

\citet{ou2004dtd} specified a language that combines static types
with dependent types.
To ensure safety, the compiler automatically inserts coercions
between dependent code and static code.
The coercions are run-time checks that ensure static code does not
crash dependent code during run-time.

\subsection{Optional Type Systems}

Optional type systems \citep{bracha2004pluggable} are an approach for
plugging static typing in dynamically typed languages through
optional type annotations that are used to perform compile-time type
checking, though not influencing the original run-time semantics of
the language.
This means that optional type systems can be unsound without affecting
language safety, because the run-time semantics should still catch
type errors independently of the static type checking.
More precisely, optional type systems are unsound if they do not catch
some type errors, but we say that languages with unsound optional type
systems are safe because the semantics of these languages catch during
run-time the type errors that the optional type system did not catch.
Although optional type systems maintain the simplicity and
flexibility of dynamically typed languages, their static type
checking cannot prove the absence of all type errors when the optional
type system is unsound.

Optional type systems have an interesting relation to the untyped
lambda calculus and the typed lambda calculus.
The semantic rules of the unytped lambda calculus are the same of the
typed lambda calculus, and the type system is actually optional and
serves only for discarding programs that may have undesired behaviors
\citep{bracha2004pluggable}.

Optional type systems make software evolution easier and
allow language designers to experiment with
\textit{pluggable type systems} \citep{bracha2004pluggable}.
Pluggable type systems are different optional type systems for
checking different aspects of the program.

According to \citet{bracha2004pluggable}, optional type systems should
not include mandatory type inference too, because mandatory type
inference may decrease the expressiveness of the type system.
Instead, optional type systems should implement type inference as
an independent mechanism that can be turned on and off according to
the programmers' needs.

Unfortunately, there is no formalization on optional type systems,
so each language ends up implementing its optional type system on
its own way.
Now we will briefly discuss some languages that provide an informal
description of their optional type systems rather than a formal one.

Strongtalk \citep{bracha1993strongtalk,bracha1996strongtalk} is
a version of Smalltalk that comes with an optional type system that
does not include any type inference.

Dart \citep{dart} and TypeScript \citep{typescript} are new
languages that are designed with an optional type system, and both
use JavaScript as their code generation target because their main
purpose is web development.
In fact, Dart is a new class-based object-oriented language with
optional type annotations while TypeScript is a strict superset of
JavaScript that provides optional type annotations and class-based
object-oriented programming.
Both Dart and TypeScript perform type inference to deduce static
types for unannotated code.

\subsection{Gradual Typing}

>>>>

\begin{verbatim}
    local Shape = { x = 0, y = 0 }
    const function Shape:new(x: number, y: number)
      local s = setmetatable({},
                             { __index = self })
      s.x = x
      s.y = y
      return shape
    end
    const function Shape:move(dx: number,
                              dy: number)
      self.x = self.x + dx
      self.y = self.y + dy
    end
    local shape1 = Shape:new(0, 5)
    local shape2: Shape = Shape:new(10, 10)
\end{verbatim}

In the last line of the example, notice how we can refer
to {\tt Shape} in the type annotation, as a shortcut
to the recursive table type that Typed Lua has assigned
to this variable.

@@

In the first line of the redefinition of {\tt new},
notice how we can call {\tt Shape}'s constructor inside
the overriden constructor. A limitation of this class
system is that the overriden constructors must be a
subtype of the original constructor.

@@

Common LISP introduced optional type annotations in the early
eighties~\citep{steele1982ocl}, but they were optimization
hints to the compiler instead of types for static checking.
These annotations were unsafe, and could crash the program
when wrong.

\citet{abadi1989dts} used tagged pairs and explicit
injection and projection operations (coercions) to embed dynamic
typing in the simply-typed lambda calculus. Dynamically-typed
values had a {\tt Dynamic} static type. \citet{thatte1990qst}
removes the necessity of explicit coercions with a system
that automatically inserts coercions and checks them for
correctness.

\textit{Soft typing}~\citep{cartwright1991soft} starts
with a dynamically-typed language, and layers a static
type system with a sophisticated global type inference
algorithm on top, to try to find errors in programs
without needing to rewrite them. In cases where an error
may or may not be present, it warns the programmer and
inserts a runtime check. One problem with the soft
typing approach was the complexity of the inferred
types, leading to errors that were difficult to understand
and fix.

\textit{Dynamic typing}~\citep{henglein1994dts} is
another approach for optimizing dynamically-typed programs.
First the program is translated to a program that uses
a {\tt Dynamic} type and explicit coercions and runtime checks,
then a static analysis removes some  of these coercions and
checks.

Instead of trying to add static checking to a dynamic
language, \citet{findler2002chf} enhances
the dynamic checks with the possibility of {\em contracts}
that give assertions about the input and output of (possibly
higher-order) functions. In case of higher-order functions,
the actual failing check can be far away from the actual
source of the error, so contracts can also add {\em blame
annotations} to values as a way to trace failures back to
the source.

Strongtalk \citep{bracha1993strongtalk,bracha1996strongtalk} is
an optionally-typed version of Smalltalk. It has a
polymorphic structural type system that programmers can use
to annotate Smalltalk programs, but type annotations can be
left out; unnanotated terms are dynamically typed, and can
be cast to any static type. The interaction of the dynamic
type with the rest of the type system is unsound, so
Strongtalk uses the dynamically-checked semantics of
Smalltalk when executing programs, even if the programs
are statically typed.

Pluggable type systems~\citep{bracha2004pluggable} generalize
the idea of Strongtalk, to have type systems that can be
layered on top of a dynamic language without influencing
its runtime semantics. These systems can be unsound in
themselves, or in their interaction with the dynamically
typed part of the language, without sacrificing runtime
safety, as the semantics of the language catch any
runtime errors caused by an unsound type system.

Dart~\citep{dart} and TypeScript~\citep{typescript} are
two recent examples of languages with optional type systems
in this style. Dart is an object-oriented language with
a semantics that is similar to Smalltalk's, while TypeScript
is an object-oriented extension of JavaScript. Dart has a
nominal type system, while TypeScript has a structural one,
but both type systems are designed on purpose with unsound
parts (such as covariant arrays in case of Dart, and
covariant function return types in case of TypeScript) to
increase programmer convenience. The interaction of
statically and dynamically typed code is also unsound.

...........HERE........
